Cluster Analysis: Identifying Parkinson's Disease Subtypes

### 2017-02-28 narrative/abbreviated.tex
Make abbreviated narrative the actual narrative

Section~\ref{sec:longitudinal} on longitudinal analysis; section~\ref{sec:nms30} on

@@

First, I used the clustering with the original 9 nms symptoms, but plotted the original 30 symptoms
instead. The reason for this was to treat the clustering of the 9 symptoms as a form of
dimensionality reduction, since we know each symptom of nms\{1-30\} is colinear with several other
symptoms, and they form the broader nms\_d\{1-9\} symptoms. The results are in
Figure~\ref{fig:d9to30}

@@

I ran PCA on the dataset with solely the 30 nonmotor symptoms. A plot of various metrics for
determining the optimal number of eigenvectors is in Figure~\ref{fig:nms30only-eigs}. Most metrics agree
Nthat the optimal number of eigenvalues is around 6, but notice how one component has by far the
highest eigenvalue; this component can be interpreted as a general ``nonmotor severity'' measure,
indicating that much of the general nonmotor symptomatic expression falls along the same dimension.

@@

\begin{figure*}[p]
  \centering
  \includegraphics[width=0.6\linewidth]{nms30m-k4-heatmap-improved.png}
  \caption{Heatmap for $k$-means clustering, $k = 4$, with the 30 nonmotor symptoms. Notice that
  subtype 2 is a ``Depression-Dominant'' subtype.} \label{fig:nms30m-k4-heatmap-improved}
\end{figure*}

\begin{table*}[b]
  \centering
  \caption{Mean symptom scores for the centers of the $k$-means clustering where $k = 4$. In
  this case, a Depression-Dominant subtype has been identified.}
  \label{tab:nms30_k4}
  \begin{tabular}{l|l|l|l|l}
cluster & 1 & 2 & 3 & 4 \\
\hline
d1-1-lightheaded & 2.08 & 2.43 & 0.55 & 5.18 \\
d1-2-fainting & 0.17 & 0.53 & 0.13 & 2.45 \\
d2-3-drowsiness & 2.51 & 2.77 & 0.97 & 7 \\
d2-4-fatigue & 4.84 & 6.29 & 1.3 & 7.31 \\
d2-5-insomnia & 3.16 & 5.46 & 1.2 & 4.22 \\
d2-6-rls & 1.88 & 2.31 & 0.66 & 3.88 \\
d3-7-loss\_interest & 1.04 & \textbf{6.08} & 0.41 & 4.59 \\
d3-8-loss\_activities & 1.7 & \textbf{7.14} & 0.7 & 5.37 \\
d3-9-anxiety & 1.97 & \textbf{6.63} & 0.91 & 4.57 \\
d3-10-depression & 2.35 & \textbf{7.81} & 0.81 & 4.29 \\
d3-11-flat\_affect & 0.91 & \textbf{4.95} & 0.39 & 2.71 \\
d3-12-loss\_pleasure & 1.09 & \textbf{6.82} & 0.36 & 3.43 \\
d4-13-hallucination & 0.58 & 0.82 & 0.2 & 4.63 \\
d4-14-delusion & 0.16 & 1.49 & 0.11 & 3.14 \\
d4-15-diplopia & 0.56 & 1.01 & 0.16 & 4.39 \\
d5-16-loss\_concentration & 2.47 & 4.02 & 0.91 & 6.55 \\
d5-17-forget\_explicit & 2.18 & 3.77 & 0.86 & 6.69 \\
d5-18-forget\_implicit & 1.76 & 3.12 & 0.68 & 6.65 \\
d6-19-drooling & 3 & 2.99 & 0.67 & 6.1 \\
d6-20-swallowing & 1.74 & 1.24 & 0.37 & 4.53 \\
d6-21-constipation & 3.31 & 3.63 & 1.65 & 6.96 \\
d7-22-urinary\_urgency & 3.84 & 3.49 & 0.99 & 7.8 \\
d7-23-urinary\_frequency & 3.79 & 3.98 & 0.94 & 6.51 \\
d7-24-nocturia & 5.13 & 4.99 & 1.67 & 7.65 \\
d8-25-sex\_drive & 2.34 & 4.46 & 0.76 & 5.16 \\
d8-26-sex\_dysfunction & 2.39 & 3.21 & 0.79 & 4.61 \\
d9-27-unexplained\_pain & 2.98 & 2.48 & 0.7 & 4.24 \\
d9-28-gust\_olfact & 3.18 & 3.57 & 1.49 & 4.73 \\
d9-29-weight\_change & 1.86 & 2.54 & 0.88 & 3.92 \\
d9-30-sweating & 2.77 & 2.2 & 0.68 & 3.18 \\
\hline
tremor & 2.59 & 2.12 & 2.53 & 4.1 \\
bradykin & 2.83 & 2.76 & 1.98 & 3.86 \\
rigidity & 2.61 & 2.42 & 1.87 & 3.61 \\
axial & 4.22 & 4.4 & 2.17 & 7.24 \\
  \end{tabular}
\end{table*}

@@

I created decision trees for the $k$-means $k = 4$ clustering, located in
Figure~\ref{fig:nms30m-dtree-all}

\begin{figure*}[p]
  \centering
  \includegraphics[width=\linewidth]{nms30m-dtree-all.pdf}
  \caption{Pruned decision tree for one vs all, clustered on nms\{1-30\}}
  \label{fig:nms30m-dtree-all}
\end{figure*}

\section{Preliminary Conclusions}

@@

>>>>

Section~\ref{sec:longitudinal} on longitudinal analysis; section~\ref{sec:nms30} on Extended
Nonmotor Symptoms; Section~\ref{ssec:newconc} on new conclusions from these sections.

@@

First, I used the clustering with the original 9 nms symptoms, but plotted the original 30 symptoms
instead. The reason for this was to treat the clustering of the 9 symptoms as a form of
dimensionality reduction, since we know each symptom of nms\{1-30\} is colinear with several other
symptoms, and they form the broader nms\_d\{1-9\} symptoms. The results are in
Figure~\ref{fig:d9to30}.

@@

I ran PCA on the dataset with solely the 30 nonmotor symptoms. A plot of various metrics for
determining the optimal number of eigenvectors is in Figure~\ref{fig:nms30only-eigs}. Most metrics agree
that the optimal number of eigenvalues is around 6, but notice how one component has by far the
highest eigenvalue; this component can be interpreted as a general ``nonmotor severity'' measure,
indicating that much of the general nonmotor symptomatic expression falls along the same dimension.

@@

Using one-way ANOVA for multiple means, only one variable had no significant interaction with the
cluster ($p > 0.05$): pdonset.

% Post-hoc analysis using Tukey's HSD is available in
% Table~\ref{tab:tukeyhsd30}. For brevity, only insignificant differences are included.

ibegin{figure*}[p]
  \centering
  \includegraphics[width=0.6\linewidth]{nms30m-k4-heatmap-improved.png}
  \caption{Heatmap for $k$-means clustering, $k = 4$, with the 30 nonmotor symptoms. Notice that
  subtype 2 is a ``Depression-Dominant'' subtype.} \label{fig:nms30m-k4-heatmap-improved}
\end{figure*}

\begin{table*}[p]
  \centering
  \caption{Mean symptom scores for the centers of the $k$-means clustering on nms\{1-30\} where $k
= 4$. In this case, a Depression-Dominant subtype has been identified.}
  \label{tab:nms30_k4}
  \begin{tabular}{l|l|l|l|l}
cluster & 1 & 2 & 3 & 4 \\
\hline
d1-1-lightheaded & 2.08 & 2.43 & 0.55 & 5.18 \\
d1-2-fainting & 0.17 & 0.53 & 0.13 & 2.45 \\
d2-3-drowsiness & 2.51 & 2.77 & 0.97 & 7 \\
d2-4-fatigue & 4.84 & 6.29 & 1.3 & 7.31 \\
d2-5-insomnia & 3.16 & 5.46 & 1.2 & 4.22 \\
d2-6-rls & 1.88 & 2.31 & 0.66 & 3.88 \\
d3-7-loss\_interest & 1.04 & \textbf{6.08} & 0.41 & 4.59 \\
d3-8-loss\_activities & 1.7 & \textbf{7.14} & 0.7 & 5.37 \\
d3-9-anxiety & 1.97 & \textbf{6.63} & 0.91 & 4.57 \\
d3-10-depression & 2.35 & \textbf{7.81} & 0.81 & 4.29 \\
d3-11-flat\_affect & 0.91 & \textbf{4.95} & 0.39 & 2.71 \\
d3-12-loss\_pleasure & 1.09 & \textbf{6.82} & 0.36 & 3.43 \\
d4-13-hallucination & 0.58 & 0.82 & 0.2 & 4.63 \\
d4-14-delusion & 0.16 & 1.49 & 0.11 & 3.14 \\
d4-15-diplopia & 0.56 & 1.01 & 0.16 & 4.39 \\
d5-16-loss\_concentration & 2.47 & 4.02 & 0.91 & 6.55 \\
d5-17-forget\_explicit & 2.18 & 3.77 & 0.86 & 6.69 \\
d5-18-forget\_implicit & 1.76 & 3.12 & 0.68 & 6.65 \\
d6-19-drooling & 3 & 2.99 & 0.67 & 6.1 \\
d6-20-swallowing & 1.74 & 1.24 & 0.37 & 4.53 \\
d6-21-constipation & 3.31 & 3.63 & 1.65 & 6.96 \\
d7-22-urinary\_urgency & 3.84 & 3.49 & 0.99 & 7.8 \\
d7-23-urinary\_frequency & 3.79 & 3.98 & 0.94 & 6.51 \\
d7-24-nocturia & 5.13 & 4.99 & 1.67 & 7.65 \\
d8-25-sex\_drive & 2.34 & 4.46 & 0.76 & 5.16 \\
d8-26-sex\_dysfunction & 2.39 & 3.21 & 0.79 & 4.61 \\
d9-27-unexplained\_pain & 2.98 & 2.48 & 0.7 & 4.24 \\
d9-28-gust\_olfact & 3.18 & 3.57 & 1.49 & 4.73 \\
d9-29-weight\_change & 1.86 & 2.54 & 0.88 & 3.92 \\
d9-30-sweating & 2.77 & 2.2 & 0.68 & 3.18 \\
\hline
tremor & 2.59 & 2.12 & 2.53 & 4.1 \\
bradykin & 2.83 & 2.76 & 1.98 & 3.86 \\
rigidity & 2.61 & 2.42 & 1.87 & 3.61 \\
axial & 4.22 & 4.4 & 2.17 & 7.24 \\
  \end{tabular}
\end{table*}

% \begin{table*}[p]
  % \centering
  % \caption{Tukey's HSD for $k$-means clustering on nms\{1-30\}, $k = 4$. For brevity, only insignificant differences ($p > 0.05$) are included.}
  % \label{tab:tukeyhsd30}
  % \begin{tabular}{l|l|l}
    % Variable & Cluster Relation & p \\
    % \hline
% d1-1-lightheaded & 3-2 & 0.572 \\
% \hline
% d1-2-fainting & 3-1 & 0.974 \\
% \hline
% d2-3-drowsiness & 3-2 & 0.814 \\
% \hline
% d2-4-fatigue & 4-2 & 0.213 \\
% \hline
% d2-5-insomnia & 4-2 & 0.143 \\
% & 4-3 & 0.168 \\
% \hline
% d2-6-rls & 3-2 & 0.504 \\
% \hline
% d4-13-hallucination & 3-2 & 0.553 \\
% \hline
% d4-14-delusion & 3-1 & 0.977 \\
% \hline
% d4-15-diplopia & 3-2 & 0.0843 \\
% \hline
% d6-19-drooling & 3-2 & 1.00 \\
% \hline
% d6-20-swallowing & 3-2 & 0.167 \\
% \hline
% d6-21-constipation & 3-2 & 0.877 \\
% \hline
% d7-22-urinary\_urgency & 3-2 & 0.758 \\
% \hline
% d7-23-urinary\_frequency & 3-2 & 0.952 \\
% \hline
% d7-24-nocturia & 3-2  & 0.983 \\
% \hline
% d8-25-sex\_drive & 4-2 & 0.575 \\
% \hline
% d8-26-sex\_dysfunction & 3-2 & 0.147 \\
% & 4-2 & 0.0635 \\
% \hline
% d9-27-unexplained\_pain & 3-2 & 0.513 \\
% \hline
% d9-28-gust\_olfact & 3-2 & 0.794 \\
% & 4-2 & 0.220 \\
% \hline
% d9-29-weight\_change & 3-2 & 0.105 \\
% \hline
% d9-30-sweating & 3-2 & 0.359 \\
% & 4-2 & 0.215 \\
% & 4-3 & 0.797 \\
% \hline
% age & 2-1 & 0.352 \\
% & 3-2 & 0.570 \\
% \hline
% sex & 3-1 & 0.443 \\
% & 4-1 & 0.991 \\
% & 3-2 & 0.373 \\
% & 4-2 & 0.436 \\
% & 4-3 & 0.969 \\
% \hline
% durat\_pd & 2-1 & 0.067 \\
 % & 3-2 & 0.436 \\
% \hline
% cisitot & 3-2 & 0.428 \\
% \hline
% tremor & 2-1 & 0.484 \\
% & 3-1 & 0.990 \\
% & 3-2 & 0.429 \\
% \hline
% bradykin & 3-2 & 0.972 \\
% \hline
% rigidity & 3-2 & 0.610  \\
% \hline
% axial & 3-2 & 0.912
  % \end{tabular}
% \end{table*}

@@

I created decision trees for the $k$-means $k = 4$ clustering, located in
Figure~\ref{fig:nms30m-dtree-all}.

\begin{figure*}[p]
  \centering
  \includegraphics[width=\linewidth]{nms30m-dtree-all.pdf}
  \caption{Pruned decision tree, clustered on nms\{1-30\}}
  \label{fig:nms30m-dtree-all}
\end{figure*}

One vs all decision trees are located in
Figures~\ref{fig:nms30m-dtree-1va}, \ref{fig:nms30m-dtree-2va}, \ref{fig:nms30m-dtree-3va},
and \ref{fig:nms30m-dtree-4va}.

\begin{figure*}[p]
  \centering
  \includegraphics[width=0.6\linewidth]{nms30m-dtree-1va.pdf}
  \caption{Pruned 1 vs all decision tree.}
  \label{fig:nms30m-dtree-1va}
\end{figure*}

\begin{figure*}[p]
  \centering
  \includegraphics[width=0.6\linewidth]{nms30m-dtree-2va.pdf}
  \caption{Unpruned 2 vs all decision tree. (Pruned was too simple).}
  \label{fig:nms30m-dtree-2va}
\end{figure*}

\begin{figure*}[p]
  \centering
  \includegraphics[width=\linewidth]{nms30m-dtree-3va.pdf}
  \caption{Pruned 3 vs all decision tree.}
  \label{fig:nms30m-dtree-3va}
\end{figure*}

\begin{figure*}[p]
  \centering
  \includegraphics[width=0.6\linewidth]{nms30m-dtree-4va.pdf}
  \caption{Unpruned 4 vs all decision tree. (Pruned was too simple).}
  \label{fig:nms30m-dtree-4va}
\end{figure*}
\section{Preliminary Conclusions}

@@

\subsection{New conclusions}
\label{ssec:newconc}

First, the longitudinal analysis gives more insight into the clusters found when clustering on
nms\_d\{1-9\}. According to Figure~\ref{fig:longcorr}, most symptoms are highly correlated, with PD
duration, but notably, mood/cognition symptoms (nms9, nms10, nms12) and tremor are not correlated
highly with PD duration. The differences in disease progression can be seen by the corresponding
graphs, Figures~\ref{fig:nms9-multi} and~\ref{fig:nms10-multi}. In both graphs, what is interesting
is that Subtype 2 (Nonmotor-Dominant) starts at higher scores for nms\_9 and nms\_10, thus
indicating that these patients' subtype is can be determined early in PD duration from depressive
symptom score. Similarly, when examining Subtype 3 (Motor-dominant) in
Figure~\ref{fig:tremor-multi}, the mean tremor score is substantially higher from PD onset.
Interestingly, Subtype 4 (Severe) generally starts at lower tremor and motor scores during disease
onset (Figure~\ref{fig:cisitot-multi}), but then rises sharply, exceeding other Subtypes. More evidence
that tremor is a unique motor symptom is located in Figure~\ref{fig:nms30m-colhc}, where it is the most
distant symptom from all other symptoms.

When examining all 30 symptoms, more evidence is given that the previously-discovered Subtype
2 (Nonmotor-Dominant) may be primarily characterized by high depressive symptoms. By graphing the
30 subtypes attached to the original clustering, as in Figure~\ref{fig:d9to30}, the mean scores of
nms8, nms9, and nms10 for Subtype 2 are substantially higher than Subtypes 1 and 3.

Indeed, PCA on the 30 nonmotor symptoms identifies the second-most prominent component as a general
mood/cognition component, and $k$-means clustering on the 30 symptoms only
(Figure~\ref{fig:nms30m-k4-heatmap-improved}) divides the 1000 patients into four slightly
different groups, a mild, average, depression-dominant, and severe group.

The Gaussian mixture model identified in Figure~\ref{fig:nms30m-mclust-heatmap-improved} fragments
the previously-discovered clusters into more groups. Here, a wide variety of specialized subtypes
of PD are displayed, including insomnia, urinary, motor, nonmotor, and depression-dominant groups,
as well as the expected mild, average, and severe subtypes. It is likely that the previous analysis
with only nms\_d\{1-9\} combined most of those specialized groups into the more general
nonmotor-dominant Subtype 2.

It's intuitive that a Depression-Dominant group emerges when clustering on nms\{1-30\}, since
domain 3 consists of 5 separate measures. Thus, any high expression of depressive symptoms is
magnified in clustering, since the symptoms are highly similar (Figure~\ref{fig:nms30m-colhc}) and
treated with equal weight. Once again reinforcing what was discovered previously, depressive
symptoms have been shown to be very important in determining subtypes of PD.
